---
title: "Assignment3"
author: "RF"
date: "3/6/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Multivariate linear models

In this assignment you will investigate the relation between different aspects of IQ and symptom severity in ASD (as measured by ADOS). The assignment is meant to make you practice on linear models, multiple predictors and interactions.

The data you will use is the outcome of the psychological testing of the children you analyzed in methods 3.
Data: https://www.dropbox.com/s/hom2qnmp9hyfwmu/Assignment3Data.csv?dl=0
The data consists of ChildID, gender, age, diagnosis (ASD), symptom severity (ADOS), several aspects of IQ (NonVerbal/Perceptual,Verbal,Social) and the visit at which the tests were administered. The questions should be answered by focusing on visit 1.
```{r}
setwd("C:/Users/JARVIS/Desktop/Uni/Semester 4/Computational modelling/Portfolio-3")

#libraries
pacman::p_load(readr,groupdata2,ggplot2,tidyverse,data.table, rethinking)

data <- read.csv("Assignment3Data.csv")
```

The questions you will have to answer are:

1. Assess the relation between symptom severity and IQ (focus on visit 1 and children with ASD) and report the model, a plot of the model, a couple of lines describing the quality of the model and interpreting the results. P.S. Should you scale? P.P.S. Don't forget to motivate your priors. P.P.P.S. At least one plot for results and a plot for quality of each model (here and in the next questions) would be appreciated.
1.1. Verbal IQ and ADOS
   
    ADOSi ~ Normal(mui,sigma)   [likelihood]  ADOSi - the symptom severity for individual i
    mui = alpha + betaQ *Qi     [linear model] Qi - Individual i's IQ
    alpha ~ Normal(0,1)         [alpha prior] 
    beta ~ Normal(0,1)          [beta prior]
    sigma ~ Uniform(0,1)        [sigma prior] 

```{r}
#focus in visit 1, children with asd
d = data %>%
  #Only choose the first visit and children with asd
  filter(Visit == 1& ASD==1)


#assess the relation between symptom severity (ADOS) and IQ
#1.1. verbal iq 
#standardize predictor - good habit
d$VerbalIQ.s <- (d$VerbalIQ-mean(d$VerbalIQ))/sd(d$VerbalIQ)
d$NonVerbalIQ.s <- (d$NonVerbalIQ-mean(d$NonVerbalIQ))/sd(d$NonVerbalIQ)
d$ADOS.s <- (d$ADOS-mean(d$ADOS))/sd(d$ADOS)

#justifying priors? Using old methods
#VIQ
# summary(glm(ADOS.s~VerbalIQ.s, family=gaussian, d)) 
# MuMIn::r.squaredGLMM(glm(ADOS.s~VerbalIQ.s, family=gaussian, d))
# ggplot(d, aes(ADOS.s,VerbalIQ.s))+
#   geom_smooth()+
#   ggtitle("ADOS depending on VerbalIQ")+
#   theme_classic()
#   
# #nonvIQ
# summary(glm(ADOS.s~NonVerbalIQ.s, family=gaussian, d)) 
# MuMIn::r.squaredGLMM(glm(ADOS.s~NonVerbalIQ.s, family=gaussian, d))
# ggplot(d, aes(ADOS.s,NonVerbalIQ.s))+
#   geom_smooth()+
#   ggtitle("ADOS depending on NonVerbalIQ")+
#   theme_classic()



#fit the model
m1<- map(
  alist(
    ADOS.s~dnorm(mu,sigma),
    mu<- a + b *VerbalIQ.s, #linear model
    a ~ dnorm(0,1), #expect that average IQ (when the standardized IQ = 0) to result in an average standardized ADOS (around  0) and you expect the standard deviations to be 1. 2 standard deviations should explain 95% of the variance, being liberal here
    b ~ dnorm(0,1), #might be not strict enough..
    sigma~dunif(0,2) #if ados is not standardized, it's mean value is around 13, 5 is a pretty liberal prior. but when standardized the mean value is 0 and 
  ),
   data= d)

precis(m1)
plot(precis(m1))
```
  alpha:
  beta for Verbal IQ:
  sigma: 
  
```{r}
#compute percentile interval of mean
#define sequence of weights to compute predictions for! 
#these values will be on the horizontal axis
VIQ.seq <- seq(from=-3, to = 3, length.out = 50) 

#simulate ados scored, the default number of simulations is 1000, I increased it here to make simulation variance smoother
sim.ados <- sim(m1, data=list(VerbalIQ.s=VIQ.seq), n =1e4)

#summarize the simulated heights
#find  89% posterior prediction interval of observable heights, across the values of weight in weight.seq
ados.PI <- apply(sim.ados,2, PI, prob=0.89)

#use link to compute mu
#for each sample from posterior
#and for each value of verbal iq in VIQ.seq
mu <- link(m1, data = data.frame(VerbalIQ.s=VIQ.seq))
mu.mean <- apply(mu, 2, mean)
mu.HPDI <- apply( mu , 2 , HPDI , prob=0.89 )


#visualize it
#show the first 100 values in the distribution of mu at each weight value  
#use type="n" to hide raw data
plot(ADOS.s ~ VerbalIQ.s, data=d, col= rangi2)
#plot the MAP line, aka the mean mu for each verbal iq value
lines(VIQ.seq, mu.mean)
#plot a shaded region for 89%  PI/ draw HPDI region for line
shade(mu.HPDI, VIQ.seq)
#plot a shaded region for 89% HPDI #draw PI region for simulated ados scores
shade(ados.PI, VIQ.seq)
mtext("89% prediction interval for ADOS score, as a fucntion of Verbal IQ.")

mu<-link(m1)
mu.mean<-apply(mu, 2, mean)
mu.HPDI<-apply( mu , 2 , HPDI , prob=0.89 )

#plotting
plot( mu.mean ~ d$ADOS.s, col=rangi2 , ylim=range(mu.HPDI) ,
xlab="Observed ADOS" , ylab="Predicted ADOS" )
mtext("Verbal IQ: Predicted ADOS scores plotted against observed ADOS scores")
abline( a = 0 , b = 1, lty = 2)
for ( i in 1:nrow(d) )
lines( rep(d$ADOS.s[i],2) , c(mu.HPDI[1,i],mu.HPDI[2,i]) ,
col=rangi2 )


```
  Look up how to interpret standardized scales..
  Interpret this plot here.
  

1.2. Non Verbal IQ and ADOS
```{r}
#1.2. nonverbal iq 
#standardize predictor - good habit
d$NonVerbalIQ.s <- (d$NonVerbalIQ-mean(d$NonVerbalIQ))/sd(d$NonVerbalIQ)

#fit the model
m2<- map(
  alist(
    ADOS.s~dnorm(mu,sigma),
    mu<- a + bNQ *NonVerbalIQ.s, #linear model
    a ~ dnorm(0,1), #expect that average IQ (when the standardized IQ = 0) to result in an average ADOS (around  13) and you expect the standard deviations to be 1. 2 standard deviations should explain 95% of the variance, being liberal here
    bNQ ~ dnorm(0,1), #might be not strict enough..
    sigma~dunif(0,2) #ados is not standardized, it's mean value is around 13, 5 is a pretty liberal prior
  ),
   data= d)

precis(m2)
plot(precis(m2))
```
  alpha:
  beta:
  sigma:

```{r}
#compute percentile interval of mean
#define sequence of weights to compute predictions for! 
#these values will be on the horizontal axis
NVIQ.seq <- seq(from=-3, to = 3, length.out = 50) 

#simulate ados scored, the default number of simulations is 1000, I increased it here to make simulation variance smoother
sim.ados <- sim(m2, data=list(NonVerbalIQ.s=NVIQ.seq), n =1e4)

#summarize the simulated heights
#find  89% posterior prediction interval of observable heights, across the values of weight in weight.seq
ados.PI <- apply(sim.ados,2, PI, prob=0.89)

#use link to compute mu
#for each sample from posterior
#and for each value of verbal iq in VIQ.seq
mu <- link(m2, data = data.frame(NonVerbalIQ.s=NVIQ.seq))
mu.mean <- apply(mu, 2, mean)
mu.HPDI <- apply( mu , 2 , HPDI , prob=0.89 )


#visualize it
#show the first 100 values in the distribution of mu at each weight value  
#use type="n" to hide raw data
plot(ADOS.s ~ NonVerbalIQ.s, data=d, col= rangi2)
#plot the MAP line, aka the mean mu for each verbal iq value
lines(NVIQ.seq, mu.mean)
#plot a shaded region for 89%  PI/ draw HPDI region for line
shade(mu.HPDI, NVIQ.seq)
#plot a shaded region for 89% HPDI #draw PI region for simulated ados scores
shade(ados.PI, NVIQ.seq)
mtext("89% prediction interval for ADOS score, as a fucntion of NonVerbal IQ.")

mu<-link(m2)
mu.mean <- apply(mu, 2, mean)
mu.HPDI <- apply( mu , 2 , HPDI , prob=0.89 )

#plotting
plot( mu.mean ~ d$ADOS.s, col=rangi2 , ylim=range(mu.HPDI) ,
xlab="Observed ADOS" , ylab="Predicted ADOS" )
mtext("NonVerbal IQ: Predicted ADOS scores plotted against observed ADOS scores")
abline( a = 0 , b = 1, lty = 2)
for ( i in 1:nrow(d) )
lines( rep(d$ADOS.s[i],2) , c(mu.HPDI[1,i],mu.HPDI[2,i]) ,
col=rangi2 )
```
  Standardized scale again.
  Interpret!
  

1.3. Social IQ and ADOS
```{r}
#1.3. social iq 
#standardize predictor - good habit
d$SocialIQ.s <- (d$SocialIQ-mean(d$SocialIQ))/sd(d$SocialIQ)

#fit the model
m3<- map(
  alist(
    ADOS.s~dnorm(mu,sigma),
    mu<- a + bSQ *SocialIQ.s, #linear model
    a ~ dnorm(0,1), #expect that average IQ (when the standardized IQ = 0) to result in an average ADOS (around  13) and you expect the standard deviations to be 1. 2 standard deviations should explain 95% of the variance, being liberal here
    bSQ ~ dnorm(0,1), #might be not strict enough..
    sigma~dunif(0,2) #ados is not standardized, it's mean value is around 13, 5 is a pretty liberal prior
  ),
   data= d)

precis(m3)
plot(precis(m3))
```
  alpha:
  beta:
  sigma:
  
```{r}
#compute percentile interval of mean
#define sequence of weights to compute predictions for! 
#these values will be on the horizontal axis
SIQ.seq <- seq(from=-3, to = 3, length.out = 50) 

#simulate ados scored, the default number of simulations is 1000, I increased it here to make simulation variance smoother
sim.ados <- sim(m3, data=list(SocialIQ.s=SIQ.seq), n =1e4)

#summarize the simulated heights
#find  89% posterior prediction interval of observable heights, across the values of weight in weight.seq
ados.PI <- apply(sim.ados,2, PI, prob=0.89)

#use link to compute mu
#for each sample from posterior
#and for each value of verbal iq in VIQ.seq
mu <- link(m3, data = data.frame(SocialIQ.s=SIQ.seq))
mu.PI <- apply(mu, 2, PI)
mu.mean <- apply(mu, 2, mean)

#visualize it
#show the first 100 values in the distribution of mu at each weight value  
#use type="n" to hide raw data
plot(ADOS.s ~ SocialIQ.s, data=d, col= rangi2)
#plot the MAP line, aka the mean mu for each verbal iq value
lines(SIQ.seq, mu.mean)
#plot a shaded region for 89%  PI
shade(mu.PI, SIQ.seq)
#plot a shaded region for 89%  PI
shade(ados.PI, SIQ.seq)
mtext("89% prediction interval for ADOS score, as a fucntion of Social IQ.")

mu<-link(m3)
mu.mean <- apply(mu, 2, mean)
mu.HPDI <- apply( mu , 2 , HPDI , prob=0.89 )

#plotting
plot( mu.mean ~ d$ADOS.s, col=rangi2 , ylim=range(mu.HPDI) ,
xlab="Observed ADOS" , ylab="Predicted ADOS" )
mtext("Social IQ: Predicted ADOS scores plotted against observed ADOS scores")
abline( a = 0 , b = 1, lty = 2)
for ( i in 1:nrow(d) )
lines( rep(d$ADOS.s[i],2) , c(mu.HPDI[1,i],mu.HPDI[2,i]) ,
col=rangi2 )

```
  Standardized scale.
  
  Try GGPLOT!!!!!
  

2. Do the different aspects of IQ account for different portions of the variance in ADOS? 
```{r}
cor(select(d, VerbalIQ.s, NonVerbalIQ.s, SocialIQ.s))

####### Thanks Pernille and Kenneth!
# this unshared variance explains any of the variance in ADOS and anaylising the betas seems to indicate that this unshared variance might explain some of the variance in ADOS 
resid_df = data.frame(residtype = c(rep("SIQ", length(d$ADOS.s)), rep("VIQ", length(d$ADOS.s)), rep("NVIQ", length(d$ADOS.s))),
                   resid = c(lm(SocialIQ.s~VerbalIQ.s+NonVerbalIQ.s, d)$residuals, v_resid = lm(VerbalIQ.s~SocialIQ.s+NonVerbalIQ.s, d)$residuals, nv_resid = lm(NonVerbalIQ.s~VerbalIQ.s+SocialIQ.s, d)$residuals),
                   ADOS.s = rep(d$ADOS.s, 3)
           )

ggplot(resid_df, aes(x = ADOS.s, y = resid, color=residtype))+
  geom_smooth(method = lm, aes(fill=residtype)) + 
  geom_jitter() +
  theme_minimal()+
  ggtitle("Residuals of IQ correlation models plotted against each other")


```


2.1. Does it make sense to have all IQ measures in the same model? First write a few lines
answering the question and motivating your answer, including a discussion as to what happens when you put all of them in the same model. Then build a model following your answer. If your answer is "no", you are not free, you still have to answer: are there alternative ways of answering the question?
ABOVE!!!!!!!!!

2.2. Build the model, assess its quality, write a few lines interpreting the results.
```{r}
#fit the model
MultiModel<- map(
  alist(
    ADOS.s~dnorm(mu,sigma),
    mu <- a+bv*VerbalIQ.s + bnv*NonVerbalIQ.s+bs*SocialIQ.s, #multivariate linear model
    a ~ dnorm(0,1), 
    bv ~ dnorm(0,1), 
    bnv ~ dnorm(0,1),
    bs ~ dnorm(0,1),
    sigma~dunif(0,2) 
  ),
   data= d)

precis(MultiModel, corr = T)
plot(precis(MultiModel))



```

I want to try smth:
```{r}
#call link without specifying new data, so it uses original data to find the mean of variables
mu <- link(MultiModel)

#summarize samples across cases
mu.mean <- apply(mu, 2, mean)
mu.PI <- apply(mu, 2, PI)

#these values will be on the horizontal axis
M.seq <- seq(from=-3, to = 3, length.out = 50) 



# #visualize it
# #show the first 100 values in the distribution of mu at each weight value  
# #use type="n" to hide raw data
# plot(ADOS.s ~ VerbalIQ.s+ NonVerbalIQ.s+SocialIQ.s, data=d, col= rangi2)
# #plot the MAP line, aka the mean mu for each verbal iq value
# lines(M.seq, mu.mean)
# #plot a shaded region for 89%  PI
# shade(mu.PI, M.seq)
# #plot a shaded region for 89%  PI
# shade(multi.PI, M.seq)
# mtext("89% prediction interval for ADOS score, as a fucntion of all IQ variables.")

#plot the simulated observations (predictions) against observed
plot(mu.mean ~ d$ADOS.s, col=rangi2, ylim=range(mu.PI),
     xlab="Observed ADOS scores", ylab="Predicted ADOS scores")
mtext("All IQ Variables: Predicted ADOS scores plotted against observed ADOS scores")
#add a line to show perfect prediction 
abline(a=0, b=1, lty=2)
#add line segments for the confidence interval of each prediction
#loop over all cases 
for (i in 1:nrow(d))
  #draw the line segment
  lines(rep(d$ADOS.s[i],2), c(mu.PI[1,i],mu.PI[2,i]),
        col=rangi2)
```



3. Let's now include also the TD children. Does it make sense to ask whether IQ and ADOS are related? Motivate your answer. In any case, if you wanted to build a model to answer that question, which model would you build? Run the model, assess its quality, write a few lines interpreting the results.
```{r}
#focus in visit 1, children with asd and td too
dt = data %>%
  #Only choose the first visit
  filter(Visit == 1)

dt <- na.omit(dt)

rescalelist = c("NonVerbalIQ", "VerbalIQ", "SocialIQ", "ADOS") #list of variables which should be rescaled

dt.s = dt[, colnames(dt) %in% rescalelist] %>% #select rows to rescale
  lapply(. , function(x) scale(x, center = mean(x, na.rm =T), scale = sd(x, na.rm = T))) %>%  #adding mean and sd manually to add the na.rm = T to removes NA's
  cbind(. , dt[,! colnames(dt) %in% rescalelist]) #bind with remaining rows

# d$VerbalIQ.s <- (d$VerbalIQ-mean(d$VerbalIQ))/sd(d$VerbalIQ)
# d$NonVerbalIQ.s <- (d$NonVerbalIQ-mean(d$NonVerbalIQ))/sd(d$NonVerbalIQ)
# d$ADOS.s <- (d$ADOS-mean(d$ADOS))/sd(d$ADOS)

#fit the model
MultiModel<- map(
  alist(
    ADOS~dnorm(mu,sigma),
    mu <- a+bv*VerbalIQ + bnv*NonVerbalIQ+bs*SocialIQ, #multivariate linear model
    a ~ dnorm(0,1), 
    bv ~ dnorm(0,1), 
    bnv ~ dnorm(0,1),
    bs ~ dnorm(0,1),
    sigma~dunif(0,2) 
  ),
   data= dt.s)

precis(MultiModel, corr = T)
plot(precis(MultiModel))


############# TD model
mu <- link( MultiModel )
# summarize samples across cases
mu.mean <- apply( mu , 2 , mean )
mu.HPDI <- apply( mu , 2 , HPDI )
# simulate observations
# again no new data, so uses original data
ADOS.sim <- sim( MultiModel , n=1e4 )
ADOS.HPDI <- apply( ADOS.sim , 2 , HPDI )

#plotting
plot( mu.mean ~ dt.s$ADOS , col=rangi2 , ylim=range(mu.HPDI) ,
xlab="Observed ADOS" , ylab="Predicted ADOS" )
mtext(" TD + ASD |Predicted ADOS scores plotted against observed ADOS scores")
abline( a = 0 , b = 1, lty = 2)
for ( i in 1:nrow(dt.s) )
lines( rep(dt.s$ADOS[i],2) , c(mu.HPDI[1,i],mu.HPDI[2,i]) ,
col=rangi2 )

```

So that's the model
Now.. answering questions:
```{r}
#looking up if iq and ados are related
summary(dt$ADOS)
summary(dt$ADOS[dt$ASD == 0])
sum(dt$ADOS[dt$ASD == 0] == 0)/length(dt$ADOS[dt$ASD == 0])
sum(dt$ADOS[dt$ASD == 1] == 0)/length(dt$ADOS[dt$ASD == 1])
sd(dt$ADOS[dt$ASD == 0])
sd(dt$ADOS[dt$ASD == 1])
mean(dt$ADOS[dt$ASD == 0])

dt$ADOS<- as.numeric(dt$ADOS)
dt$ASD<- as.factor(dt$ASD)
ggplot(dt, aes(ADOS,na.rm= T, color=ASD))+
  geom_density()+
  theme_minimal()+
  scale_colour_discrete(name = "Diagnosis", labels = c("TD", "ASD"))+
  ggtitle("Distribution of ADOS scores across TDs and ASDs")

ggplot(dt, aes(ADOS,na.rm= T))+
  geom_density(color="green")+
  theme_minimal()+
  #scale_colour_discrete(name = "Diagnosis", labels = c("TD", "ASD"))+
  ggtitle("Distribution of ADOS scores across ASDs and TDs together")

ggplot(dt, aes(ADOS, VerbalIQ,color=ASD))+
  geom_smooth()+
  geom_jitter()+
  theme_minimal()+
  scale_colour_discrete(name = "Diagnosis", labels = c("TD", "ASD"))+
  ggtitle("Verbal IQ depending on ADOS scores in TDs and ASDs")

ggplot(dt, aes(ADOS, NonVerbalIQ,color=ASD))+
  geom_smooth()+
  geom_jitter()+
  theme_minimal()+
  scale_colour_discrete(name = "Diagnosis", labels = c("TD", "ASD"))+
  ggtitle("NonVerbal IQ depending on ADOS scores in TDs and ASDs")

ggplot(dt, aes(ADOS, SocialIQ,color=ASD))+
  geom_smooth()+
  geom_jitter()+
  theme_minimal()+
  scale_colour_discrete(name = "Diagnosis", labels = c("TD", "ASD"))+
  ggtitle("Social IQ depending on ADOS scores in TDs and ASDs")

ggplot(dt, aes(ASD, SocialIQ, color=ASD))+
  geom_violin()+
  theme_minimal()+
  scale_colour_discrete(name = "Diagnosis", labels = c("TD", "ASD"))+
  ggtitle("Social IQ depending on diagnosis")

dt<- na.omit(dt)

m <- map( #adding TD to the model
  alist(
    ADOS ~ dnorm(mu, sigma),
    mu <- a+bs*SocialIQ+bv*VerbalIQ + bnv*NonVerbalIQ + bd * ASD,
    a ~ dnorm(0, 1), 
    bs ~ dnorm(0, 1),
    bv ~ dnorm(0, 1),
    bnv ~ dnorm(0, 1),
    bd ~ dnorm(1, 0.5),
    sigma ~ dunif(0,2)
  ), data = dt.s)

precis(m)

#call link without specifying new data, so it uses original data to find the mean of variables
mu <- link(m)

#summarize samples across cases
mu.mean <- apply(mu, 2, mean)
mu.PI <- apply(mu, 2, PI)

#these values will be on the horizontal axis
M.seq <- seq(from=-3, to = 3, length.out = 50) 

#plot the simulated observations (predictions) against observed
plot(mu.mean ~ dt$ADOS, col=rangi2, ylim=range(mu.PI),
     xlab="Observed ADOS scores", ylab="Predicted ADOS scores")
mtext("IQ Variables and Diagnosis: Predicted ADOS scores plotted against observed ADOS scores")
#add a line to show perfect prediction 
abline(a=0, b=1, lty=2)
#add line segments for the confidence interval of each prediction
#loop over all cases 
for (i in 1:nrow(dt))
  #draw the line segment
  lines(rep(dt$ADOS[i],2), c(mu.PI[1,i],mu.PI[2,i]),
        col=rangi2)

```


4. Let's discuss contents:
4.1. You have three scores for IQ, do they show shared variance? Is that the same in TD and ASD? What does that tell us about IQ?
4.2. You have explored the relation between IQ and ADOS. How do you explain that relation from a cognitive perspective? N.B. You can present alternative hypotheses.

5. Bonus questions: Including measurement errors. 
5.1. Let's assume that ADOS has a measurement error of 1. How would you model that in any of the previous models? 
5.2. We know that IQ has an estimated measurement error of 2.12. How would you include that? 


